---
title: "PanCan_2013"
author: "Russel Sutherland"
date: "13th May 2015"
output: html_document
---


This document shows the exploratory analyses I conducted on the PanCancer data.This data comes from the SAGE synapse repository (accession syn1710680) (https://www.synapse.org/#!Synapse:syn1710680).

I have principal component analyses and correlations / wilcoxon /kruskal wallis tests to establish associations with clinical features.
The MDS plots allow us to see any potential technical effects present in the data.

```{r, echo=FALSE}
library(scales)
col1<-c("#33a02c", "#e31a1c", "#ff7f00", "#6a3d9a", "#ffff99", "#b15928")
palette(col1)
# the stratification object
load("/Users/rds/Dropbox/PhD/TCGA_pipeline/MDS_data/PanCanStrat.RData")
#the mutationmatrix
load("/Users/rds/Dropbox/PhD/TCGA_pipeline/MDS_data/PanCan_mutationMatrixLogical2.RData")
#the ISOMDS object
load("/Users/rds/Dropbox/PhD/TCGA_pipeline/MDS_data/PanCan_ISOMDS.RData")
#the silent mutation rate data
load("/Users/rds/Dropbox/PhD/TCGA_pipeline/MDS_data/PanCan_silentMutationsPerIndiv.RData")######run this and save it when I get home

```

The number of colon and rectum samples
```{r, echo=TRUE}
 table(PanCanstrat$tumourType[,2])
```

The breakdown of cancer type by sequencer type and test for over-representation
```{r,echo=FALSE}
table(PanCanstrat$sequenceCenter[,2], PanCanstrat$tumourType[,2])
chisq.test(PanCanstrat$sequenceCenter[,2], PanCanstrat$tumourType[,2])
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.




Principal component analysis conducted on the Binary Mutation Matrix.
```{r, echo=FALSE}

#TCGA nature principal component analysis
PanCan_prcomp<-prcomp(t(PanCan_mutationMatrixLogical2), scale=TRUE)
#TCGA_prcomp_mutTable<-prcomp(t(mutationTable), scale=TRUE)

screeplot(PanCan_prcomp, type="lines",col="dark grey", main="TCGA June PCA screeplot")
```

Variance from the first 10 principal components
```{r,echo=FALSE}
summary(PanCan_prcomp)[[6]][,1:10]
```


##Correlations of features with the first principal component
```{r, echo=FALSE}

#correlation of PC1 with metadata
cor.test(PanCan_prcomp$x[,1], PanCan_silentMutationsPerIndiv, method = "spearman")# correlation with mutation rate
```
relationship of age with PC1
```{r, echo=FALSE}

#correlation of PC1 with metadata
cor.test(PanCan_prcomp$x[,1], PanCanstrat$age_at_initial_pathologic_diagnosis[,2], method = "spearman")# correlation with age
```




sequencer relationship to PC1
```{r, echo=FALSE}
kruskal.test(PanCan_prcomp$x[,1]~as.factor(PanCanstrat$sequenceCenter[,2]))
# correlation with sequence center
by(PanCan_prcomp$x[,1],PanCanstrat$sequenceCenter[,2], FUN=mean)
#post hoc comparisons
pairwise.wilcox.test(PanCan_prcomp$x[,1], PanCanstrat$sequenceCenter[,2], p.adj="bonferroni",exact=F)
prcompseq<-data.frame(PC1=PanCan_prcomp$x[,1],seqCenter=PanCanstrat$sequenceCenter[,2])
prcompseq<-prcompseq[PanCanstrat$sequenceCenter[,2]!="[Not Available]",]
library(coin)
#effect size comparison between BCM and Washington University
pcseqtest<-wilcox_test(PC1~factor(seqCenter), data=prcompseq)
pcseqtest
```

effectsize
```{r, echo=FALSE}
pcseqtest@statistic@teststatistic/sqrt(220)

```
gender relationship to PC1
```{r, echo=FALSE}
#cor.test(TCGA_prcomp$x[,1], TCGA_stratification$age_at_initial_pathologic_diagnosis[,2])# correlation with age
wilcox.test(PanCan_prcomp$x[,1]~ PanCanstrat$gender[,2], conf.int = TRUE)#  do the genders has different pc1 scores?
by(PanCan_prcomp$x[,1],PanCanstrat$gender[,2], FUN=mean)
```
cancer type relationship with PC1
```{r, echo=FALSE}
wilcox.test(PanCan_prcomp$x[,1]~ PanCanstrat$tumourType[,2], conf.int = TRUE)# correlation with mutation gender
by(PanCan_prcomp$x[,1],PanCanstrat$tumourType[,2], FUN=mean)

```


## Multi Dimensional Scaling Plots

```{r, echo=FALSE}

MDSplot<- function(mdsPoints,smplclrs,hyperMutIndex){
  par(mar=c(6,6,6,6))
  par(xpd=TRUE)
  plot(mdsPoints$points[,1],mdsPoints$points[,2], pch = 19, col=alpha(smplclrs[,1],0.5), cex=0.7, cex.lab=0.7, cex.main = 0.7, cex.axis=0.7, xlab="dimension 1", ylab="dimension 2")
  points(mdsPoints$points[hyperMutIndex,1], mdsPoints$points[hyperMutIndex,2],pch=1, cex = 0.9)
  #legend(min(v$points[,1])+0.01, max(v$points[,2]), bty="n", c("SOLiD", "IlluminaHiSeq","not available","hypermutated"), cex=3, pch=c(19,19,19,1),col=c(col1[c(2,1,3)], "black"))
  legend(min(mdsPoints$points[,1])+0.01, max(mdsPoints$points[,2]),bty="n", c(unique(smplclrs[,2]),"hypermutated"), cex=0.7,pt.cex=0.7, pch=c(rep(19, length(unique(smplclrs[,2]))),1),col=c(seq_along(1:length(unique(smplclrs[,1]))),"black"))
}
```

#### sequencer MDS plot

```{r, fig.width=8, fig.height=8, echo=FALSE}

MDSplot(PanCan_ISOMDS,PanCanstrat$sequenceCenter,which(PanCanstrat$hypermutated[,2]==TRUE))

```

#### microsatellite instability MDS plot

```{r, fig.width=8, fig.height=8, echo=FALSE}
MDSplot(PanCan_ISOMDS,PanCanstrat$mononucleotide_and_dinucleotide_marker_panel_analysis_status,which(PanCanstrat$hypermutated[,2]==TRUE))


```