---
title: "TCGA June (Nature publication) data exploratory analysis"
author: "Russel Sutherland"
date: "8 May 2015"
output: html_document
---

This document shows the exploratory analyses I conducted on the TCGA June 2012 data.This data comes from the supplementary data from the comprehensive molecular characterisation of colorectal cancer paper (http://www.nature.com/nature/journal/v487/n7407/full/nature11252.html).

I have principal component analyses and correlations / wilcoxon tests to establish associations with clinical features.
The MDS plots allow us to see any potential technical effects present in the data.

```{r, echo=FALSE}
library(scales)
col1<-c("#33a02c", "#e31a1c", "#ff7f00", "#6a3d9a", "#ffff99", "#b15928")
palette(col1)
# the stratification object
load("/Users/rds/Dropbox/PhD/TCGA_pipeline/MDS_data/TCGA_Nature_strat.RData")
#the mutationmatrix
load("/Users/rds/Dropbox/PhD/TCGA_pipeline/MDS_data/TCGA_Nature_mutationMatrixLogical2.RData")
#the ISOMDS object
load("/Users/rds/Dropbox/PhD/TCGA_pipeline/MDS_data/TCGA_Nature_ISOMDS.RData")
#the silent mutation rate data
load("/Users/rds/Dropbox/PhD/TCGA_pipeline/MDS_data/TCGA_Nature_silentMutationsPerIndiv.RData")

```

The number of colon and rectum samples
```{r, echo=TRUE}
 table(TCGA_Nature_strat$cancer[,2])
```

The breakdown of cancer type by sequencer type and test for over-representation
```{r,echo=FALSE}
table(TCGA_Nature_strat$seqTech[,2], TCGA_Nature_strat$cancer[,2])
chisq.test(TCGA_Nature_strat$seqTech[,2], TCGA_Nature_strat$cancer[,2])
```
Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.




Principal component analysis conducted on the Binary Mutation Matrix.
```{r, echo=FALSE}

#TCGA nature principal component analysis
TCGA_Nature_prcomp<-prcomp(t(TCGA_Nature_mutationMatrixLogical2), scale=TRUE)
#TCGA_prcomp_mutTable<-prcomp(t(mutationTable), scale=TRUE)

screeplot(TCGA_Nature_prcomp, type="lines",col="dark grey", main="TCGA June PCA screeplot")
```

Variance from the first 10 principal components
```{r,echo=FALSE}
summary(TCGA_Nature_prcomp)[[6]][,1:10]
```


##Correlations of features with the first principal component
```{r, echo=FALSE}

#correlation of PC1 with metadata
cor.test(TCGA_Nature_prcomp$x[,1], TCGA_Nature_silentMutationsPerIndiv, method = "spearman")# correlation with mutation rate
```
relationship of age with PC1
```{r, echo=FALSE}

#correlation of PC1 with metadata
cor.test(TCGA_Nature_prcomp$x[,1], TCGA_Nature_strat$age_at_initial_pathologic_diagnosis[,2], method = "spearman")# correlation with age
```




sequencer relationship to PC1
```{r, echo=FALSE}
kruskal.test(TCGA_Nature_prcomp$x[,1]~ as.factor(TCGA_Nature_strat$seqTech[,2]))# correlation with sequence center
by(TCGA_Nature_prcomp$x[,1],TCGA_Nature_strat$seqTech[,2], FUN=mean)
```
gender relationship to PC1
```{r, echo=FALSE}
#cor.test(TCGA_prcomp$x[,1], TCGA_stratification$age_at_initial_pathologic_diagnosis[,2])# correlation with age
wilcox.test(TCGA_Nature_prcomp$x[,1]~ TCGA_Nature_strat$gender[,2])#  do the genders has different pc1 scores?
by(TCGA_Nature_prcomp$x[,1],TCGA_Nature_strat$gender[,2], FUN=mean)
```
cancer type relationship with PC1
```{r, echo=FALSE}
wilcox.test(TCGA_Nature_prcomp$x[,1]~ TCGA_Nature_strat$cancer[,2])# correlation with mutation gender
by(TCGA_Nature_prcomp$x[,1],TCGA_Nature_strat$cancer[,2], FUN=mean)

```


## Multi Dimensional Scaling Plots

```{r, echo=FALSE}

MDSplot<- function(mdsPoints,smplclrs,hyperMutIndex){
  par(mar=c(6,6,6,6))
  par(xpd=TRUE)
  plot(mdsPoints$points[,1],mdsPoints$points[,2], pch = 19, col=alpha(smplclrs[,1],0.5), cex=0.7, cex.lab=0.7, cex.main = 0.7, cex.axis=0.7, xlab="dimension 1", ylab="dimension 2")
  points(mdsPoints$points[hyperMutIndex,1], mdsPoints$points[hyperMutIndex,2],pch=1, cex = 0.9)
  #legend(min(v$points[,1])+0.01, max(v$points[,2]), bty="n", c("SOLiD", "IlluminaHiSeq","not available","hypermutated"), cex=3, pch=c(19,19,19,1),col=c(col1[c(2,1,3)], "black"))
  legend(min(mdsPoints$points[,1])+0.01, max(mdsPoints$points[,2]),bty="n", c(unique(smplclrs[,2]),"hypermutated"), cex=0.7,pt.cex=0.7, pch=c(rep(19, length(unique(smplclrs[,2]))),1),col=c(seq_along(1:length(unique(smplclrs[,1]))),"black"))
}
```

#### sequencer MDS plot

```{r, fig.width=8, fig.height=8, echo=FALSE}

MDSplot(TCGA_Nature_ISOMDS,TCGA_Nature_strat$seqTech,which(TCGA_Nature_strat$hypermutated[,2]==TRUE))

```

#### microsatellite instability MDS plot

```{r, fig.width=8, fig.height=8, echo=FALSE}
MDSplot(TCGA_Nature_ISOMDS,TCGA_Nature_strat$MSI_status,which(TCGA_Nature_strat$hypermutated[,2]==TRUE))


```